Distortion corrected kernel density estimate on Riemannian manifolds

### 2022-11-06 paper/paper.tex
Update PDF, table electricityoutliers hash for color codes to be solved

\caption{\label{tab:fourdhdrtable}Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.}
\centering
\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{UMAP} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
\midrule
>99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
\bottomrule
\end{tabular}
\end{table}

\autoref{tab:fourdhdrtable} further illustrates the percentage of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions. Compared with KDE, the percentages are always higher when distortion correction methods are used. The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\). This could be due to the severe distortion usually induced by the UMAP algorithm.

@@

For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}. This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions. \textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds. Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a 2 dimensional embedding for kernel density estimation and anomaly detection. To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used. We use the highest density regions plot to visualize the density estimates. However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.

Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here. For each household, a discrete approximation to the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found. For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn matric algorithm. The bandwidth parameter \(r\) is set as 1 in Equation \eqref{eq:denestimator}.

@@

{\centering \includegraphics[width=0.9\linewidth]{figures/kde_2densities_raw_electricity}

@@

\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.}\label{fig:electricityoutliers}
\end{figure}

\autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
The typical households in the top row are similar except that there are a few spikes for household \(1472\) (selected as most typical when using KDE without distortion correction). As for the anomalies, distortion corrected KDE tends to capture instance of unusual electricity demand where usage is very low with some sudden spikes such as ID 3243 or households with a very high baseline electricity usage only across a sub-period of time such as ID 2838. In contrast, KDE without distortion correction can select households that do not exhibit anomalous behavior at all, such as household 5146 and the selection of such houesholds as anomalies may be an artefact of distortion induced by a dimension reduction algorithm.

@@

{\centering \includegraphics[width=0.9\linewidth]{figures/hdr_electricity_hdrbox_3id_7dow}

@@

\caption{Quantile region plots of electricity demand against the time of week for one typical household 1321 and two anomalies, 3243 and 2838. The quantile regions displayed in the plot are 99.0\%, 95.0\%, 75\%, and 50\%. Household 1321 has a repeated period-of-the-week electricity usage pattern slightly different for weekdays and weekends.}\label{fig:electricityhdrsfixed}
\end{figure}

@@

{\centering \includegraphics[width=0.9\linewidth]{figures/electricity_hdrbox_3id_7dow}

@@

Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde} which are respectively selected when distortion correction is and is not applied. Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
This repeated pattern in a week window is also obvious for the typical household ID 1321 from DC-KDE. As for the households selected when distortion correction is applied, the middle row in \autoref{fig:electricityhdrsfixed} shows a household with small spikes at noon but only on Tuesday and Sunday. The anomalies detected without distortion correction show an increasing electricity demand across the day of the week. These findings show the difference between anomalies detected with different kernel density estimation methods.

@@

As an empirical example, we explore the distributions of \(3,639\) households and \(336\) time periods of the week in the Irish smart meter data. Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE. Without the ground truth density, we compare both density estimates by looking at the distributions of the most typical households with the highest densities and the anomalous households with the top two lowest densities. Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.

There are several open questions to be explored. The first involves the selection of bandwidth parameters for kernel density estimation, which has been explored in many KDE-related literatures. The second one is related to the quality of the embedding from dimension reduction methods. Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitive relationship between embedding quality and density estimation accuracy is not immediate clear. The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell wherther the distortion are corrected in the right way. The density estimates on the edges of the whole data structure could alos be explored because most outer area points tend to be detected as outliers. However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.

>>>>

\caption{\label{tab:fourdhdrtable}Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings.}
\centering
\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{UMAP} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
\midrule
>99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
\bottomrule
\end{tabular}
\end{table}

\autoref{tab:fourdhdrtable} further illustrates the proportions of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions. Compared with KDE, the proportions are always higher when distortion correction methods are used. The proportions with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\). This could be due to the severe distortion usually induced by the UMAP algorithm.

@@

For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}. This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions. \textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds. Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a \(2\)-dimensional embedding for kernel density estimation and anomaly detection. To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used. We use the highest density regions plot to visualize the density estimates. However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.

Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here. For each household, a discrete approximation to the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found. For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm. The number of nearest neighbors in manifold learning \(K\) is set as \(100\) and the bandwidth paramter \(r=180\) to match the distances between the distributions of households.

@@

{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_5levels_outliers_comparison_5ml_r180}

@@

\caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel). The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityembedding}
\end{figure}
\begin{table}

\caption{\label{tab:electricityoutliers}Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding. The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE. The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.}
\centering
\begin{tabu} to \linewidth {>{}c>{}c>{}c>{}c>{}c>{}c|>{}c>{}c>{}c>{}c>{}c}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{5}{c}{DC-KDE} & \multicolumn{5}{c}{KDE} \\
\cmidrule(l{3pt}r{3pt}){2-6} \cmidrule(l{3pt}r{3pt}){7-11}
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ISOMAP} & \multicolumn{1}{c}{LLE} & \multicolumn{1}{c}{Laplacian Eigenmaps} & \multicolumn{1}{c}{t-SNE} & \multicolumn{1}{c}{UMAP} & \multicolumn{1}{c}{ISOMAP} & \multicolumn{1}{c}{LLE} & \multicolumn{1}{c}{Laplacian Eigenmaps} & \multicolumn{1}{c}{t-SNE} & \multicolumn{1}{c}{UMAP} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7} \cmidrule(l{3pt}r{3pt}){8-8} \cmidrule(l{3pt}r{3pt}){9-9} \cmidrule(l{3pt}r{3pt}){10-10} \cmidrule(l{3pt}r{3pt}){11-11}
1 & \textcolor{black}{6317} & \textcolor{black}{7049} & \textcolor{black}{3161} & \textcolor{black}{4826} & \textcolor{black}{4546} & \textcolor{black}{6317} & \textcolor{black}{1881} & \textcolor{black}{6317} & \textcolor{black}{4190} & \textcolor{black}{6533}\\
2 & \textcolor{0}{1353} & \textcolor[HTML]{08306B}{\textbf{4546}} & \textcolor[HTML]{08306B}{\textbf{5269}} & \textcolor{0}{1069} & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{4884} & \textcolor{0}{6199} & \textcolor{0}{6643} & \textcolor{0}{6817} & \textcolor{0}{5899}\\
3 & \textcolor{0}{4884} & \textcolor[HTML]{08306B}{\textbf{1881}} & \textcolor{0}{4875} & \textcolor{0}{4546} & \textcolor[HTML]{08306B}{\textbf{1069}} & \textcolor{0}{5092} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor{0}{6764} & \textcolor{0}{6492} & \textcolor{0}{6345}\\
4 & \textcolor{0}{4546} & \textcolor{0}{6952} & \textcolor{0}{7300} & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{4862} & \textcolor{0}{1353} & \textcolor[HTML]{6BAED6}{\textbf{4233}} & \textcolor{0}{1058} & \textcolor{0}{2309} & \textcolor{0}{1884}\\
5 & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{6838} & \textcolor{0}{4880} & \textcolor[HTML]{08306B}{\textbf{1533}} & \textcolor{0}{3639} & \textcolor{0}{5580} & \textcolor{0}{7049} & \textcolor{0}{5105} & \textcolor{0}{2546} & \textcolor{0}{3278}\\
\addlinespace
6 & \textcolor[HTML]{08306B}{\textbf{1366}} & \textcolor{0}{3874} & \textcolor{0}{5599} & \textcolor[HTML]{6BAED6}{\textbf{4265}} & \textcolor{0}{1964} & \textcolor{0}{5594} & \textcolor[HTML]{08306B}{\textbf{1821}} & \textcolor{0}{6577} & \textcolor{0}{2907} & \textcolor{0}{5463}\\
7 & \textcolor{0}{3161} & \textcolor{0}{1069} & \textcolor{0}{3222} & \textcolor{0}{4862} & \textcolor[HTML]{6BAED6}{\textbf{2249}} & \textcolor{0}{2576} & \textcolor{0}{4546} & \textcolor{0}{3880} & \textcolor{0}{4536} & \textcolor{0}{5504}\\
8 & \textcolor[HTML]{08306B}{\textbf{1074}} & \textcolor{0}{2249} & \textcolor{0}{1776} & \textcolor{0}{3639} & \textcolor{0}{6199} & \textcolor{0}{1366} & \textcolor[HTML]{08306B}{\textbf{3431}} & \textcolor{0}{5358} & \textcolor{0}{6898} & \textcolor{0}{1782}\\
9 & \textcolor{0}{6681} & \textcolor{0}{1964} & \textcolor{0}{1161} & \textcolor{0}{5546} & \textcolor[HTML]{6BAED6}{\textbf{3892}} & \textcolor{0}{6681} & \textcolor{0}{1569} & \textcolor{0}{1366} & \textcolor{0}{4128} & \textcolor{0}{5965}\\
10 & \textcolor{0}{1701} & \textcolor[HTML]{6BAED6}{\textbf{1569}} & \textcolor{0}{3655} & \textcolor{0}{4896} & \textcolor{0}{6209} & \textcolor{0}{5291} & \textcolor{0}{1964} & \textcolor{0}{5545} & \textcolor{0}{7368} & \textcolor{0}{6081}\\
\addlinespace
11 & \textcolor{0}{3136} & \textcolor{0}{1533} & \textcolor{0}{6169} & \textcolor{0}{4555} & \textcolor{0}{6952} & \textcolor{0}{3161} & \textcolor[HTML]{6BAED6}{\textbf{6568}} & \textcolor{0}{2160} & \textcolor{0}{2232} & \textcolor{0}{4894}\\
12 & \textcolor{0}{1881} & \textcolor[HTML]{6BAED6}{\textbf{4555}} & \textcolor{0}{4830} & \textcolor{0}{6199} & \textcolor{0}{4896} & \textcolor[HTML]{08306B}{\textbf{1701}} & \textcolor{0}{7300} & \textcolor{0}{6585} & \textcolor{0}{6273} & \textcolor{0}{5937}\\
13 & \textcolor{0}{6785} & \textcolor{0}{6526} & \textcolor{0}{2284} & \textcolor[HTML]{6BAED6}{\textbf{6952}} & \textcolor{0}{6838} & \textcolor{0}{4749} & \textcolor{0}{7438} & \textcolor{0}{5824} & \textcolor{0}{1557} & \textcolor{0}{5522}\\
14 & \textcolor{0}{5092} & \textcolor{0}{3639} & \textcolor{0}{6526} & \textcolor{0}{6838} & \textcolor{0}{1533} & \textcolor{0}{6699} & \textcolor{0}{4401} & \textcolor{0}{4932} & \textcolor{0}{3500} & \textcolor{0}{5580}\\
15 & \textcolor{0}{5269} & \textcolor{0}{6199} & \textcolor{0}{1694} & \textcolor{0}{1964} & \textcolor[HTML]{6BAED6}{\textbf{1569}} & \textcolor{0}{5504} & \textcolor{0}{3434} & \textcolor{0}{6410} & \textcolor{0}{1932} & \textcolor{0}{6902}\\
\addlinespace
16 & \textcolor{0}{2576} & \textcolor[HTML]{6BAED6}{\textbf{5531}} & \textcolor{0}{3431} & \textcolor[HTML]{6BAED6}{\textbf{7104}} & \textcolor{0}{3213} & \textcolor{0}{3170} & \textcolor{0}{2249} & \textcolor{0}{3023} & \textcolor{0}{3119} & \textcolor{0}{7243}\\
17 & \textcolor{0}{1262} & \textcolor{0}{4875} & \textcolor{0}{2523} & \textcolor{0}{3892} & \textcolor{0}{4401} & \textcolor{0}{5899} & \textcolor{0}{3213} & \textcolor{0}{1497} & \textcolor{0}{3809} & \textcolor{0}{6699}\\
18 & \textcolor{0}{5966} & \textcolor{0}{5356} & \textcolor{0}{5356} & \textcolor{0}{3161} & \textcolor{0}{6766} & \textcolor{0}{2658} & \textcolor{0}{3161} & \textcolor{0}{7293} & \textcolor{0}{6108} & \textcolor{0}{5549}\\
19 & \textcolor{0}{5937} & \textcolor[HTML]{6BAED6}{\textbf{3892}} & \textcolor[HTML]{6BAED6}{\textbf{2233}} & \textcolor[HTML]{08306B}{\textbf{6799}} & \textcolor{0}{4265} & \textcolor{0}{5189} & \textcolor[HTML]{08306B}{\textbf{1533}} & \textcolor{0}{5580} & \textcolor{0}{4516} & \textcolor{0}{5048}\\
20 & \textcolor{0}{5356} & \textcolor{0}{4896} & \textcolor{0}{5125} & \textcolor{0}{6209} & \textcolor{0}{6799} & \textcolor{0}{2669} & \textcolor[HTML]{6BAED6}{\textbf{4555}} & \textcolor{0}{6224} & \textcolor{0}{5937} & \textcolor{0}{1923}\\
\bottomrule
\end{tabu}
\end{table}

\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE. There are six density regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\), \(99\%\) and \(>99\%\), and darker colors indicate lower density regions. The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities. First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings. When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE and UMAP with DC-KDE compared with the KDE. As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identifies as anomalies. The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times. Again, KDE with distortion correction methods find more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases. Without distortion correction, the density estimates are different not only in scaling, but also shows inconsistency in the rank of density estimates.

\caption{\label{tab:electricityhdrpercent}Proportion of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabu} to \linewidth {>{\raggedright}X|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}c>{\centering}X}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{t-SNE} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
\multicolumn{1}{c}{HDRs} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7} \cmidrule(l{3pt}r{3pt}){8-8} \cmidrule(l{3pt}r{3pt}){9-9}
>99\% HDR & \cellcolor[HTML]{E6E6E6}{0.189} & 0.243 & \cellcolor[HTML]{E6E6E6}{0.189} & 0.027 & \cellcolor[HTML]{E6E6E6}{0.622} & 0.000 & \cellcolor[HTML]{E6E6E6}{0.865} & 0.108\\
99\% HDR & \cellcolor[HTML]{E6E6E6}{0.359} & \textbf{0.248} & \cellcolor[HTML]{E6E6E6}{\textbf{0.503}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.697}} & 0.007 & \cellcolor[HTML]{E6E6E6}{\textbf{0.628}} & 0.083\\
\bottomrule
\end{tabu}}
\end{table}
\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_rankplot_3639id_umapvs4ml_radius20_r180}

\caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel). The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityrank}
\end{figure}

Unlike the simulations where the ground truth densities are known, in the applications, we usually have little knowledge about the anomalous distributions of the households. Therefore, we assume, in this case, the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE. The results are summarized in \autoref{tab:electricityhdrpercent} similar to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) in \(>99\%\) and \(99\%\) HDRs that are also identified in different dimension reduction methods. Both DC-KDE and KDE are shown in the table. Except the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods. The improvement brought by distortion correction is much more obvious in Laplacian Eigenmaps and t-SNE when the increase in both HDRs are greater than \(0.5\).
\autoref{fig:electricityrank} presents the visualization of density ranking in log scale. The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.

@@

{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_compare4id_1typical_3anomalous}

@@

\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.}\label{fig:electricityraw}
\end{figure}

Now let us take a close look at the typical and anomalous distributions of households. According to \autoref{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration. Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}. Compared with ID 2396, the other three households show different usage patterns. ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh. As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household. The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days. It is likely that these two households are not occupied all the time during the trial and could be used as a AirBnb place or holiday place.

@@

{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous}

@@

Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityoutliers}. The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
This repeated pattern in a week window is also shown in some days of week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays. As for the bottom two households, the middle row in \autoref{fig:electricityhdrsfixed} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week. These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.

@@

As an empirical example, we explore the distributions of \(3,639\) households and \(336\) time periods of the week in the Irish smart meter data. Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE. Without the ground truth density, we assume that UMAP gives the true densities and compare both density estimation methods by looking at the distributions of the most typical households with the highest densities and three anomalous households that are identified most times. The distortion correction methods could identify the typical households with repeated time-of-the-week usage patterns, while the distributions of the outliers are anomalous in different ways.

There are several open questions to be explored. The first involves the selection of bandwidth parameters for kernel density estimation, which has been widely explored in many KDE-related literature. The second one is that the proposed density estimation method is only limited to the embedding points. Although for the downstream task of anomaly detection, only density estimated on these points are needed, it would be worth exploring the out-of-sample density estimation in the future.
The last question is related to the quality of the embedding from dimension reduction methods. Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediate clear. The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion are corrected in the right way. The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers. However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
